---
title: "Final AKW"
author: "Aung Kyaw Win"
date: "2024-12-03"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Loading the necessary libraries

```{r}
library(dplyr)
library(ggplot2)
library(rjson)
library(plotrix)
library(car)
library(data.table)
library(ggthemes)
library(lubridate)
```

Reading the data files

```{r}
#Transactions CSV File
transactions =fread("C:/Users/akw97/Downloads/archive/transactions_data.csv")
#Merchant Categories JSON File
merchant_categories = fromJSON(file = "C:/Users/akw97/Downloads/archive/mcc_codes.json" )
mcc_df <- data.frame(mcc_code = names(merchant_categories), business_type = unlist(merchant_categories), stringsAsFactors = FALSE)

#users Data CSV File
user_data = read.csv("C:/Users/akw97/Downloads/archive/users_data.csv")

#Card Data CSV File
card_data = read.csv("C:/Users/akw97/Downloads/archive/cards_data.csv")
```

Exploring the datasets

```{r}
summary(transactions)
head(transactions)

summary(user_data)
head(user_data)

summary(card_data)
head(card_data)

#Checking for missing values
colSums(is.na(transactions)) #For Transactions File

colSums(is.na(user_data)) #For User Data File

colSums(is.na(card_data)) #For  Card Data File
```


Data Wrangling Process

```{r}
#Clean Up Process For Transactions CSV File
transactions$amount = gsub("[$,-]", "", transactions$amount) #Removing the $ sign and minus sign from the "amount" column
transactions$amount = as.numeric(transactions$amount) #Converting the column into numerics
transactions$use_chip = as.factor(transactions$use_chip) #Converting the transaction type into factors

#Clean Up Process For User Data CSV File
user_data$yearly_income = gsub("[$,]", "", user_data$yearly_income) #Removing the $ sign
user_data$yearly_income = as.numeric(user_data$yearly_income) #Converting into a numeric

user_data$total_debt = gsub("[$,]", "", user_data$total_debt) #Removing the $ sign
user_data$total_debt = as.numeric(user_data$total_debt) #Converting into a numeric

head(user_data)


#Clean Up Process For Merchant Categories JSON File
mcc_df$mcc_code = as.integer(mcc_df$mcc_code) #Converting the mcc code into integers

str(mcc_df) #Rechecking to see changes were made correctly

#Clean Up Process For Card Data CSV File
card_data$credit_limit = gsub("[$,]", "", card_data$credit_limit) #Removing the $sign
card_data$credit_limit = as.numeric(card_data$credit_limit)


#Verifying changes
head(transactions$amount)
head(user_data)
head(card_data)
head(mcc_df)
```

Merging Datasets

```{r}
#Merging all the datasets by common keys
data1 = transactions %>%
  inner_join(user_data, by = c("client_id" = "id")) #Joining User Data into Transaction Data

data2 = transactions %>%
  inner_join(card_data, by = "client_id") #Joining Card Data into Transaction Data
```

```{r}
data3 = data1 %>%
  inner_join(mcc_df, by = c("mcc" = "mcc_code")) #Joining Merchant Category Into Transaction Data and User Data

data4 = user_data %>%
  inner_join(card_data, by = c("id" = "client_id")) # Joining User Data and Card Data CSV Files

data4 = data4 %>%
  mutate(credit_utilization = (total_debt / credit_limit) * 100)

data3 = na.omit(data3) #Omitting null values
head(data3) #Checking if changes were made
```

```{r}
data3 <- data3 %>%
  mutate(
    # Creating income bins based on yearly_income
    income_bin = cut(yearly_income,
                     breaks = c(0, 30000, 70000, 150000, Inf),
                     labels = c("Low", "Middle", "Upper Middle", "High"), 
                     right = FALSE), # Ensures bins are inclusive of the left side

    # Creating credit score categories based on credit_score
    credit_score_range = case_when(
      credit_score < 600 ~ "Low",
      credit_score >= 600 & credit_score < 700 ~ "Medium",
      credit_score >= 700 ~ "High",
      TRUE ~ NA_character_ # Ensures NA values are handled if any
    ),

    # Creating Debt To Income Ratio column
    debt_to_income = total_debt / yearly_income,

    # Categorizing users into age groups
    age_group = case_when(
      current_age <= 25 ~ "Young Adult",
      current_age > 25 & current_age <= 45 ~ "Adult",
      current_age > 45 ~ "Older People"
    )
  )

#Converting date to a propoer format and grouping by month
transactions1 <- transactions %>%
  mutate(transaction_month = floor_date(as.Date(date, format = "%Y-%m-%d"), "month")) %>%
  group_by(transaction_month) %>%
  summarize(total_amount = sum(amount, na.rm = TRUE))
```

Stratified Random Sampling Dataset

```{r}
set.seed(123) #To reproduce the same results

sampled_data = data3 %>%
  group_by(income_bin) %>%
  sample_frac(1300000 / nrow(data3)) %>% #Using 10% of the dataset
  ungroup()

summary(sampled_data)
```


Transactions By Type Graph

```{r}
#Summarizing the transactions by transaction type and converting into percentage
transaction_type = transactions %>%
  group_by(use_chip) %>%
  summarize(count = n ())%>%
  mutate(percentage = count / sum(count) * 100)

#Creating a Pie Chart to Visualize
ggplot(transaction_type, aes(x = "", y = percentage, fill = use_chip,)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") + #Converting the bar plot into pie chart
  labs(title = "Transactions by Type", x = "", y = "") +
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5)) +
  theme_minimal() +
  theme(legend.position = "right")
```

Transactions Count By Top Cities
```{r}
#Summarizing the transactions count by merchant city and selecting the top 10 cities
top_cities = transactions %>%
  filter(merchant_city != "ONLINE") %>% # Exclude transactions with "online" as the merchant city
  group_by(merchant_city) %>%
  summarize(transaction_count = n()) %>%
  arrange(desc(transaction_count)) %>%
  slice_head(n = 10)

#Creating a Bar plot to visualize
ggplot(top_cities, aes(x = reorder(merchant_city, transaction_count), y = transaction_count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +
  labs(title = "Top 10 Merchant Cities by Transaction Count", x = "Merchant City", y = "Transaction Count") +
  theme_economist()
```

Transactions By Business Type

```{r}
#Summarizing the total transaction amount by business type and selecting the top 5 business categories
spending_category = data3 %>%
  group_by(business_type) %>%
  summarize(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))%>%
  slice_max(total_amount, n = 5) %>%
  mutate(percentage = total_amount / sum(total_amount) * 100) #Calculating the percentage of each business type

ggplot(spending_category, aes(x = "", y = total_amount, fill = business_type)) + 
  geom_bar(stat = "identity", width = 1, color = "white", size = 0.8) +  # White borders around each slice
  coord_polar("y") +  # Convert bar chart to pie chart
  labs(title = "Spending by MCC Code Category", x = "", y = "") +  # Title and remove axis labels
  geom_text(aes(label = paste0(round(percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),  # Position percentage labels in the center
            color = "white", size = 3, fontface = "bold") +  # White bold text for readability
  scale_fill_brewer(palette = "Paired") +  # Use the Paired color palette
  theme_minimal() +  # Remove gridlines and background
  theme(
    legend.position = "right",  # Position legend to the right
    plot.title = element_text(size = 20, face = "bold", color = "darkblue", hjust = 0.5),  # Fancy title
    plot.margin = margin(30, 30, 30, 30)  # Add padding around the plot
  )
```

Average Spending By Income Bracket and Linear Regression between Yearly Income and Transaction Amount

```{r}
avg_spending = data3 %>%
  group_by(income_bin) %>%
  summarize(average_spending = mean(amount, na.rm = TRUE))

print(avg_spending)

ggplot(avg_spending, aes(x = income_bin, y = average_spending, fill = income_bin)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    title = "Average Spending by Income Bracket",
    x = "Income Bracket",
    y = "Average Spending"
  ) +
  theme_economist() +
  scale_fill_brewer(palette = "Set3")

#Linear regression to assess the relationship between yearly income and spending
lm_model <- lm(amount ~ yearly_income, data = sampled_data)
summary(lm_model)

# Print the model's coefficients
print(coef(lm_model))

# Plotting the linear regression between yearly income and transaction amount
ggplot(sampled_data, aes(x = yearly_income, y = amount)) +
  geom_point(alpha = 0.4) +  # Scatter plot of the data points
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  # Adding the regression line
  labs(
    title = "Relationship between Yearly Income and Spending",
    x = "Yearly Income",
    y = "Transaction Amount"
  ) +
  theme_economist()
```
Average Spending By Credit Score and Linear Regression Between Credit Score and Transaction Amount
```{r}
avg_spending_by_CreditScore = data3 %>%
  group_by(credit_score_range) %>%
  summarize(average_spending = mean(amount, na.rm = TRUE))

ggplot(avg_spending_by_CreditScore, aes(x = credit_score_range, y = average_spending, fill = credit_score_range)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  labs(
    title = "Average Spending by Credit Score Range",
    x = "Credit Score Range",
    y = "Average Spending"
  ) +
  theme_economist() +
  scale_fill_brewer(palette = "Set2")

#Linear regression to assess the relationship between credit score and spending amount
lm_credit_score = lm(amount~credit_score, data = sampled_data)
summary(lm_credit_score)


# Plotting the linear regression between Credit Score and transaction amount
ggplot(sampled_data, aes(x = credit_score, y = amount)) +
  geom_point(alpha = 0.4) +  # Scatter plot of the data points
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  # Adding the regression line
  labs(
    title = "Relationship between Credit Score and Spending",
    x = "Credit Score",
    y = "Transaction Amount"
  ) +
  theme_economist()
```

Average Spending By Age and Linear Regression between Current Age and Transaction Amount
```{r}
avg_spending_by_age = data3 %>%
  group_by(current_age) %>%
  summarize(average_spending = mean(amount, na.rm = TRUE))

ggplot(avg_spending_by_age, aes(x = current_age, y = average_spending))+
 geom_point(alpha = 0.4, color = "blue") + # Scatter points
  geom_smooth(method = "lm", color = "red") + # Regression line
  labs(
    title = "Relationship between Age and Spending Amount",
    x = "Current Age",
    y = "Transaction Amount"
  ) +
  theme_economist()

#Linear regression to assess the relationship between current age and spending amount
lm_age_spending = lm(amount ~ current_age, data = sampled_data)
summary(lm_age_spending)

# Plotting the linear regression between Current Age and Transaction Amount
ggplot(sampled_data, aes(x = current_age, y = amount)) +
  geom_point(alpha = 0.4) +  # Scatter plot of the data points
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  # Adding the regression line
  labs(
    title = "Relationship between Current Age and Spending",
    x = "Current Age",
    y = "Transaction Amount"
  ) +
  theme_economist()
```

Average Debt and Age and Linear Regression Between Total Debt and Age
```{r}
avg_debt_by_age = data3 %>%
  group_by(current_age) %>%
  summarize(average_debt = mean(total_debt, na.rm = TRUE))

# Plotting the relationship
ggplot(avg_debt_by_age, aes(x = current_age, y = average_debt)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "Relationship between Age and Average Debt",
    x = "Average Age",
    y = "Average Debt"
  ) +
  theme_economist()

#Linear regression to assess the relationship between current age and total debt
lm_age_debt <- lm(total_debt ~ current_age, data = user_data)
summary(lm_age_debt)

plot(lm_age_debt)

# Plotting the linear regression between Current Age and Total Debt
ggplot(sampled_data, aes(x = current_age, y = total_debt)) +
  geom_point(alpha = 0.4) +  # Scatter plot of the data points
  geom_smooth(method = "lm", color = "blue", se = FALSE) +  # Adding the regression line
  labs(
    title = "Relationship between Total Debt and Current Age",
    x = "Age",
    y = "Total Debt"
  ) +
  theme_economist()
```

Multiple Linear Regression Predicting Transaction Amount Based on Current Age, Credit Score, and Yearly Income

```{r}
lm_multiple <- lm(amount ~ current_age + credit_score + yearly_income, data = sampled_data)
summary(lm_multiple)
```

Avg Debt By Age Group

```{r}
#Creating Age Bins and Summarizing by Average Debt
avg_debt_and_age = data3 %>%
  mutate(age_bin = cut(current_age, breaks = seq(20,100,15))) %>%
  group_by(age_bin) %>%
  summarize(avg_debt = mean(total_debt, na.rm = TRUE))

#Plotting the Graph
ggplot(avg_debt_and_age, aes(x = age_bin, y = avg_debt)) +
  geom_bar(stat = "identity", fill = "lightblue") +
  labs(title = "Average Debt By Age Group", x = "Age Group", y = "Average Debt") +
  theme_economist()
```

Linear Regression Between Yearly Income and Total Debt

```{r}
#Building the linear regression model
lm_income_debt = lm(total_debt ~ yearly_income, data = user_data)
summary(lm_income_debt)

#Plotting the graph
ggplot(user_data, aes(x = yearly_income, y = total_debt)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm", color = "blue") +
  labs(title = "Relationship Between Yearly Income and Total Debt",
       x = "Yearly Income", y = "Total Debt") +
  theme_economist()
```

Time Series Analysis on Transaction Trends Over Time

```{r}
ggplot(transactions1, aes(x = transaction_month, y = total_amount)) +
  geom_line(color = "darkgreen", size = 1) +
  labs(title = "Monthly Transaction Trends", x = "Month", y = "Total Transactions") +
  theme_economist()
```

Correlation Between Credit Limit and Credit Score

```{r}
# Calculate the correlation coefficient between credit limit and credit score
correlation <- cor(data4$credit_score, data4$credit_limit, use = "complete.obs")

# Print the correlation result
print(paste("Correlation between Credit Score and Credit Limit: ", round(correlation, 2)))

# Create a scatter plot
ggplot(data4, aes(x = credit_limit, y = credit_score)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Relationship between Credit Limit and Credit Score",
    subtitle = paste("Correlation: ", round(correlation, 2)),
    x = "Credit Limit",
    y = "Credit Score"
  ) +
  theme_economist()

```

Correlation Between Income and Credit Limit

```{r}
# Step 1: Calculate the correlation between yearly income and credit limit
correlation_income_limit <- cor(data4$yearly_income, data4$credit_limit, use = "complete.obs")

#Building a simple linear regression
lm_income_creditlimit = lm(credit_limit ~ yearly_income, data = data4)
summary(lm_income_creditlimit)

#Evaluating the regression
plot(lm_income_creditlimit)

# Step 2: Create a scatter plot
ggplot(data4, aes(x = yearly_income, y = credit_limit)) +
  geom_point(alpha = 0.3, color = "black") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Relationship between Yearly Income and Credit Limit",
    subtitle = paste("Correlation: ", round(correlation_income_limit, 2)),
    x = "Yearly Income",
    y = "Credit Limit"
  ) +
  theme_economist()

```

Linear Regression Between Total_debt and Number Of Credit Cards

```{r}
# Aggregate total debt and number of credit cards
debt_by_cards <- sampled_data %>%
  group_by(num_credit_cards) %>%
  summarize(
    avg_total_debt = mean(total_debt, na.rm = TRUE)
  )
#Building a simple linear regression
lm_debt_cards = lm(avg_total_debt ~  num_credit_cards, data = debt_by_cards)
summary(lm_debt_cards)

#Evaluating the regression
plot(lm_debt_cards)

# Plot the relationship
ggplot(debt_by_cards, aes(x = num_credit_cards, y = avg_total_debt)) +
  geom_point(color = "darkblue", size = 3, alpha = 0.6) +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(
    title = "Relationship between Number of Cards and Total Debt",
    x = "Number of Credit Cards",
    y = "Average Total Debt"
  ) +
  theme_economist()

```

Correlation Of Regional Income and Average Transaction Amount

```{r}
# Calculate average transaction amount by state
state_avg_income <- sampled_data %>%
  group_by(merchant_state) %>%
  summarize(avg_transaction = mean(amount, na.rm = TRUE),
            avg_income = mean(yearly_income, na.rm = TRUE))

print(state_avg_income)

# Scatterplot of average income vs. average transaction amount
ggplot(state_avg_income, aes(x = avg_income, y = avg_transaction)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Correlation Between Regional Income and Average Transaction Amount",
    x = "Average Yearly Income",
    y = "Average Transaction Amount"
  ) +
  theme_economist()

```


