---
title: "Final_Project_TTT"
author: "Group-5"
date: "2024-12-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

These are the following libraries to answer the question *(shown below)*

```{r}
library(dplyr)
library(ggplot2)
library(jsonlite)
library(ggthemes)
```

**Load the datasets** and rename columns names in `mcc` dataset

```{r}
data <- read.csv("D:/Education/Data Science/Programming with R/FinalProject/transactions_data.csv")
#head(data)
mcc <- read_json("D:/Education/Data Science/Programming with R/FinalProject/mcc_codes.json") %>% 
  as.data.frame() %>% 
  stack() %>% 
  mutate(ind=gsub("\\X", "", ind)) %>% 
  rename(mcc=ind, transaction_type=values)
```

**Data Segmentation** in the `transaction data` for better performance & joining with `mcc` dataset

```{r}
df_selected <- data %>% select(date, amount, use_chip, merchant_city, mcc, client_id)

df_selected$mcc <- as.character(df_selected$mcc)
df_selected <- left_join(df_selected, mcc, by = "mcc")
```

**Cleaning** the `amount` column from `selected transaction data`

```{r}
df_selected$amount <- gsub("\\$", "", df_selected$amount)  # Remove dollar signs
df_selected$amount <- gsub("^-", "", df_selected$amount) 
df_selected$amount <- gsub("-", "", df_selected$amount)

df_selected$amount <- as.numeric(df_selected$amount)
```

Changing the "Chip Transaction" to "Online Transaction" for **data consistency**

```{r}
df_selected <- df_selected %>% 
  mutate(use_chip = ifelse(use_chip == "Chip Transaction", "Online Transaction", use_chip))
```

*Question: Is there a difference in spending behavior between online (chip-based) and in-store (swipe-based) transactions?*

1.  **Calculating summary statistics for each transaction type**

```{r}
chip_based_stats <- df_selected %>% filter(use_chip == "Online Transaction") %>%
  summarise(
    count = n(),
    total_amount = sum(amount, na.rm = TRUE),
    mean_amount = mean(amount, na.rm = TRUE),
    median_amount = median(amount, na.rm = TRUE),
    std_amount = sd(amount, na.rm = TRUE)
  )

swipe_based_stats <- df_selected %>% filter(use_chip == "Swipe Transaction") %>%
  summarise(
    count = n(),
    total_amount = sum(amount, na.rm = TRUE),
    mean_amount = mean(amount, na.rm = TRUE),
    median_amount = median(amount, na.rm = TRUE),
    std_amount = sd(amount, na.rm = TRUE)
  )

chip_based_stats
swipe_based_stats
```

Looking at the statistics, averagely people tend to spend 53 - 52\$ on both platforms. However, there are more chip transactions made looking at the count.

2.  **The number of transactions for each type and payment method**

```{r}
top_transactions <- df_selected %>%
  group_by(use_chip, transaction_type) %>%
  summarise(count = n()) %>%
  arrange(desc(count)) %>%
  group_by(use_chip) %>%
  slice_head(n = 5)

ggplot(top_transactions, aes(x = reorder(transaction_type, -count), y = count, fill = use_chip)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 5 Transaction Types by Payment Method", x = "Transaction Type", y = "Count", color = "Payment Method") +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 70, hjust=1, vjust=1))
```

According to the graph, most transactions happens in grocery stores and supermarkets, food stores & service stations. But noticeably, tolls and bridge fees are paid solely by online transactions & drug stores and pharmacies seems like they only accept swipe transactions.

3.  **Payment Method Based on Merchant City**

```{r}
payment_by_location <- df_selected %>%
  group_by(use_chip, merchant_city) %>%
  summarise(count = n(), .groups = 'drop') %>%
  arrange(desc(count)) %>%
  group_by(use_chip) %>%
  slice_max(count, n = 5) %>%
  ungroup() %>% 
  arrange(desc(count))

ggplot(payment_by_location, aes(x = reorder(merchant_city, -count), y = count, fill = use_chip)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Top 5 Cities by Payment Method", x = "Merchant City", y = "Total Transactions", color = "Payment Methods") +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust=1)) +
  facet_wrap(~ use_chip, scales = "free_y", ncol = 1)
```

Right off the bat, there are more online payments compared to store based payments. For store based payments, Houston has the most payments and followed by Miami and LA.

3.  **Time Series Analysis on Payment Method Overtime**

```{r}
df_selected$date <- as.Date(df_selected$date)

time_series_data <- df_selected %>%
  group_by(date, use_chip) %>%
  summarise(total_amount = sum(amount), .groups = 'drop')

ggplot(time_series_data, aes(x = date, y = total_amount, color = use_chip, group = use_chip)) +
  geom_line(size = 0.2) +
  labs(title = "Time Series Analysis of Payment Methods", x = "Date", y = "Total Amount", color = "Payment Methods") +
  theme_economist() +
  theme(axis.text.x = element_text(angle = 45, hjust=1, vjust=1))
```

Starting from 2010, there were more store based transactions and after 2015, most transactions it switched to online. One small detail is that, during 2016, both online and store based transactions dropped to zero for some reason.
